---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo = TRUE}
# libraries
library(dplyr) # my workhorse for data manipulation and piping subsetting
library(ggvis) # Hadley Wickham new plotting system
library(magrittr) # for piping %>% and %<>%
library(knitr) # kable wouldn't work without it

# using unzip to decompres the file
# if(!file.exists("./activity.csv")) else if {
#     unzip("activity.zip", exdir = ".")}
# read csv into data frame called df.activities
df_actities <- read.csv(file = "./activity.csv", header = TRUE, sep = ",",
                        stringsAsFactors = FALSE)
df_actities$date <- as.Date(df_actities$date)
```



## What is mean total number of steps taken per day?
####1. Histogram of the activites by day
```{r, echo = TRUE, results = "asis"}
### make a histogram of the steps per day
df_actities %>%
    filter(steps >= 1) %>% # does not remove NA since NA does not equal 0
    filter(steps != "NA") %>% # To remove NAs
    ggvis(x = ~date, y = ~steps) %>%
    layer_histograms(width = 1)
```

####2. Mean of the steps per day
```{r, echo = TRUE, results = "asis"}
mean_activities <- df_actities %>%
    filter(steps >= 1) %>%
    filter(steps != "NA") %>% # To remove NAs
    group_by(date) %>%
    summarise(mean(steps))
```

Table of the days and their means of steps
```{r, echo = TRUE, results = "asis"}
kable(mean_activities)
```

####3. Median of the steps per day
```{r, echo = TRUE, results = "asis"}
median_activities <- df_actities %>%
    filter(steps >= 1) %>%
    filter(steps != "NA") %>% # To remove NAs
    group_by(date) %>%
    summarise(median(steps)) %>%
    as_data_frame()
```

Table of the days and their median of steps
```{r, echo = TRUE, results = "asis"}
kable(median_activities)
```

## What is the average daily activity pattern?
This is a plot showing the average steps taken through out the day. The time is in military time with the interval being an average of steps every 5 minutes.
*Between 8:00 AM and 9:30 AM was the most active time of day*
```{r, echo = TRUE}

df_actities %>%
    select(interval, steps) %>% 
    filter(steps != "NA") %>% # To remove NAs
    group_by(interval) %>% 
    summarise_each(funs(mean)) %>%
    ggvis(x = ~interval, y = ~steps) %>% 
    layer_lines() %>%
    add_axis("x" , title = "Time of Day (Military Time") %>%
    add_axis("y", title = "Steps in 5 Minutes")

```
## 5 minute interval that contains the max number of steps
```{r, echo = TRUE}
# Create a data frame of the mean of the intervals
df_interval_mean <- df_actities %>%
    select(interval, steps) %>% 
    filter(steps != "NA") %>% # To remove NAs
    group_by(interval) %>% 
    summarise_each(funs(mean))
# row identified
df_interval_mean[which.max(df_interval_mean$steps), ]
```

# Missing Values handled

## Number of missing values
1. sum of missing values
```{r, echo = TRUE}
# sum of missing values
sum(is.na(df_actities))
```

## Creating a new data frame with the average of the intervals replacing the NA values in steps
```{r, echo = TRUE}
# Create a data frame of the mean of the intervals
df_interval_mean <- df_actities %>%
    select(interval, steps) %>% 
    filter(steps != "NA") %>% # To remove NAs
    group_by(interval) %>% 
    summarise_each(funs(mean))
# merge data_frames of both and join on interval
df_actities_fixed <- left_join(df_actities, df_interval_mean, by = "interval")
# ifelse which column is needed
df_actities_fixed$steps <- ifelse(is.na(df_actities_fixed$steps.x), df_actities_fixed$steps.y, df_actities_fixed$steps.x)
# remove the steps.x and steps.y
df_actities_fixed %<>%
    select(date, interval, steps)
```

## What is mean total number of steps taken per day?
####1. Histogram of the activites by day (NA replaced with average values for interval)
```{r, echo = TRUE, results = "asis"}
### make a histogram of the steps per day
df_actities_fixed %>%
    filter(steps >= 1) %>% # does not remove NA since NA does not equal 0
    filter(steps != "NA") %>% # To remove NAs
    ggvis(x = ~date, y = ~steps) %>%
    layer_histograms(width = 1)
```


### mean and median of steps per day
####2. Mean of the steps per day (NA replaced with average values for interval)
```{r, echo = TRUE, results = "asis"}
mean_activities <- df_actities_fixed %>%
    filter(steps >= 1) %>%
    filter(steps != "NA") %>% # To remove NAs
    group_by(date) %>%
    summarise(mean(steps))
```

Table of the days and their means of steps
```{r, echo = TRUE, results = "asis"}
kable(mean_activities)
```

####3. Median of the steps per day (NA replaced with average values for interval)
```{r, echo = TRUE, results = "asis"}
median_activities <- df_actities_fixed %>%
    filter(steps >= 1) %>%
    filter(steps != "NA") %>% # To remove NAs
    group_by(date) %>%
    summarise(median(steps)) %>%
    as_data_frame()
```

Table of the days and their median of steps
```{r, echo = TRUE, results = "asis"}
kable(median_activities)
```

## Are there differences in activity patterns between weekdays and weekends?
```{r, echo = TRUE, results = "asis"}
df_paterns <- df_actities_fixed
df_paterns$date <- as.Date(df_paterns$date)
df_paterns$day.names <- weekdays(df_actities_fixed$date)
df_paterns$day.type <- ifelse(df_paterns$day.names %in% c("Saturday", "Sunday"), "weekend", "weekday")

```

## Plots to show diffference between weekday and weekend activities
#### Weekday Graph
```{r, echo = TRUE, results = "asis"}
suppressWarnings(df_paterns %>%
    select(interval, steps, day.type) %>%
    filter(day.type == "weekday") %>% 
    group_by(interval) %>% 
    summarise_each(funs(mean)) %>%
    ggvis(x = ~interval, y = ~steps) %>% 
    layer_lines() %>%
    add_axis("x" , title = "Time of Day (Military Time") %>%
    add_axis("y", title = "Steps in 5 Minutes"))
```
1. Weekday graph shows burst of activity

2. Max time of activity was 8:00 AM - 9:30 AM

3. Low levels of activity throughout the rest of the day

#### Weekend Graph

```{r, echo = TRUE, results = "asis"}
suppressWarnings(df_paterns %>%
    select(interval, steps, day.type) %>%
    filter(day.type == "weekend") %>% 
    group_by(interval) %>% 
    summarise_each(funs(mean)) %>%
    ggvis(x = ~interval, y = ~steps) %>% 
    layer_lines() %>%
    add_axis("x" , title = "Time of Day (Military Time") %>%
    add_axis("y", title = "Steps in 5 Minutes"))
```
1. Weekend show much more activitity throughout the day

2. Max time of activity was still 8:00 AM - 9:30 AM

3. Activity stayed elevated through the day